Higher-Order unification for free!

### 2024-04-25 paper/main.tex
massive rename F -> O bject, H -> M eta, T -> Q , N -> n

\newcommand{\Ue}{\ensuremath{\UnifRel_\lambda}\xspace}
\newcommand{\nUe}{\ensuremath{\nUnifRel_\lambda}\xspace}
\newcommand{\Ee}{\ensuremath{\EqualRel_\lambda}\xspace}
\newcommand{\nEe}{\ensuremath{\nEqualRel_\lambda}\xspace}
\newcommand{\llambda}{\ensuremath{\mathcal{L}}\xspace}

@@

%\newcommand{\Fo}{\texorpdfstring{\ensuremath{\mathcal{F}_{\!o}\xspace}}{Fo}} % space non va
%\newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{H}_o}\xspace}{Ho}}
\newcommand{\Fo}{\texorpdfstring{\ensuremath{\mathcal{F}\xspace}}{F}} % space non va
\newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{H}}\xspace}{H}}

@@

\newcommand{\linkStore}{\texorpdfstring{\ensuremath{\mathbb{L}}\xspace}{L}}
\newcommand{\mapStore}{\texorpdfstring{\ensuremath{\mathbb{M}}\xspace}{M}}
\newcommand{\foUnifPb}{\ensuremath{\mathbb{P}}\xspace}
\newcommand{\hoUnifPb}{\ensuremath{\mathbb{T}}\xspace}

@@

Even if we encountered the problem working on Coq's Calculus of Inductive
Constructions, we devise
a minimal setting to ease its study. In this setting, we have
a \Fo{} language (for first order) with a rich equational
theory and a \Ho{} meta language with a simpler one.

@@

\subsection{The problem: logic-program simulation}
We represent a logic program \emph{run} in \Fo{} as
a sequence of \emph{steps} of length $\mathcal{N}$.
At each step $p$ we unify two terms, $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$, taken from the list of all unification
problems \foUnifPb.
% \footnote{If the same rule is used multiple times in a run we
% consider as many copies as needed of the terms composing the
% rules, with fresh unification variables each time. Also, we do not
% model backtracking directly, but rather the fact that a step in a run may
% fail, and in that case the entire run is said to fail. A regular logic program
% that succeeds is modeled as a many runs, with possibly overlapping starts,
% where at least one run does not end in a failure. }
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$, which is the result of the
logic program execution.
%
\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[4]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFx}[3]{\ensuremath{\frun(#1,#2) \mapsto #3}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHx}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho'}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}\vspace{2pt}\\
\runF{\foUnifPb}{\mathcal{N}}{\rho}
\deff
\runFD{\foUnifPb}{\mathcal{N}}
\end{array}
$$

@@

We simulate each run in \Fo{} with a run in \Ho as follows:
%
%Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\hoUnifPb}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
  \runH{\foUnifPb}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t,m,l) | s \in \foUnifPb, \C{s}{t}{m}{l} \}\vspace{2pt}\\
  \qquad \runHD{\hoUnifPb}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mapStore}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
\end{array}
$$

@@

\begin{proposition}[Run equivalence]\label{prop:simulation}
$\forall \foUnifPb, \forall \mathcal{N},$ if $~\foUnifPb \subseteq \llambda$
$$
  \runFx{\foUnifPb}{\mathcal{N}}{\rho} \land
  \runHx{\foUnifPb}{\mathcal{N}}{\rho'}
  \Rightarrow
  \forall s \in \foUnifPb, \rho s \Eo \rho' s
$$
\end{proposition}

@@

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of$\;$ \hrun, if $~\foUnifPb \subseteq \llambda$ we have that
$\forall p \in 1 \ldots \mathcal{N},$
$$
\stepF{\foUnifPb}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_p}{\linkStore_p}
$$
\end{proposition}
\noindent
In particular, this property guarantees that a \emph{failure} in the \Fo{} run
is matched by a failure in \Ho{} \emph{at the same step}. We consider this
property very important from a practical point of view since it guarantees
that the execution traces are strongly related, and in turn, this enables a user
to debug a logic program in \Fo{} by looking at its execution trace in
\Ho{}.

>>>>

\newcommand{\Ue}{\ensuremath{\UnifRel_m}\xspace}
\newcommand{\nUe}{\ensuremath{\nUnifRel_m}\xspace}
\newcommand{\Ee}{\ensuremath{\EqualRel_m}\xspace}
\newcommand{\nEe}{\ensuremath{\nEqualRel_m}\xspace}
\newcommand{\llambda}{\ensuremath{\mathcal{L}}\xspace}

@@

%\newcommand{\Fo}{\texorpdfstring{\ensuremath{\mathcal{F}_{\!o}\xspace}}{Fo}} % space non va
%\newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{H}_o}\xspace}{Ho}}
\newcommand{\Fo}{\texorpdfstring{\ensuremath{\mathcal{O}\xspace}}{O}} % space non va
\newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{M}}\xspace}{M}}

@@

\newcommand{\linkStore}{\texorpdfstring{\ensuremath{\mathbb{L}}\xspace}{L}}
\newcommand{\mapStore}{\texorpdfstring{\ensuremath{\mathbb{M}}\xspace}{M}}
\newcommand{\foUnifPb}{\ensuremath{\mathbb{P}}\xspace}
\newcommand{\hoUnifPb}{\ensuremath{\mathbb{Q}}\xspace}

@@

Even if we encountered the problem working on Coq's Calculus of Inductive
Constructions, we devise
a minimal setting to ease its study. In this setting, we have
a (first order) language \Fo{} with a rich equational
theory and a \Ho{} language with a simpler one.

@@

\subsection{The problem: logic-program simulation}
We represent a logic program \emph{run} in \Fo{} as
a sequence of \emph{steps} of length $n$.
At each step $p$ we unify two terms, $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$, taken from the list of all unification
problems \foUnifPb.
% \footnote{If the same rule is used multiple times in a run we
% consider as many copies as needed of the terms composing the
% rules, with fresh unification variables each time. Also, we do not
% model backtracking directly, but rather the fact that a step in a run may
% fail, and in that case the entire run is said to fail. A regular logic program
% that succeeds is modeled as a many runs, with possibly overlapping starts,
% where at least one run does not end in a failure. }
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_n$, which is the result of the
logic program execution.
%
\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{ostep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{mstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{orun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{mrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[4]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFx}[3]{\ensuremath{\frun(#1,#2) \mapsto #3}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHx}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho'}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}\vspace{2pt}\\
\runF{\foUnifPb}{n}{\rho}
\deff
\runFD{\foUnifPb}{n}
\end{array}
$$

@@

We simulate each run in \Fo{} with a run in \Ho as follows:
%
%Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\hoUnifPb}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
  \runH{\foUnifPb}{n}{\rho} \deff \vspace{2pt}\\
  \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t,m,l) | s \in \foUnifPb, \C{s}{t}{m}{l} \}\vspace{2pt}\\
  \qquad \runHD{\hoUnifPb}{n}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{n}}{\mapStore}{\linkStore_{n}}{\rho_{n}}
\end{array}
$$

@@

\begin{proposition}[Run equivalence]\label{prop:simulation}
$\forall \foUnifPb, \forall n,$ if $~\foUnifPb \subseteq \llambda$
$$
  \runFx{\foUnifPb}{n}{\rho} \land
  \runHx{\foUnifPb}{n}{\rho'}
  \Rightarrow
  \forall s \in \foUnifPb, \rho s \Eo \rho' s
$$
\end{proposition}

@@

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of$\;$ \hrun, if $~\foUnifPb \subseteq \llambda$ we have that
$\forall p \in 1 \ldots n,$
$$
\stepF{\foUnifPb}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_p}{\linkStore_p}
$$
\end{proposition}
\noindent
In particular, this property guarantees that a \emph{failure} in the \Fo{} run
is matched by a failure in \Ho{} \emph{at the same step}. We consider this
property very important from a practical point of view since it guarantees
that the execution traces are strongly related, and in turn, this enables a user
to debug a logic program in \Fo{} by looking at its execution trace in
\Ho{}.
